---
title: "Overlap_NoPower"
output: html_document
date: "2024-11-02"
---

libraries
```{r message=FALSE, warning=FALSE, include=FALSE}
library(data.table)
library(pals)
library(ggpubr)
library(rstatix)
library(ggbeeswarm)
library(tidyverse)

wd <- "/cfDNA/withTURBT/Overlap/"
wd_corr <- "/cfDNA/withTURBT/Overlap/Correlations/"
wd_driv <- "/cfDNA/withTURBT/Overlap/Drivers/"

figs <- '/Main_Figures/Figure7/raw_images/Mutation_overlap/withTURBT/'
fig_corr <- '/Main_Figures/Figure7/raw_images/Mutation_overlap/withTURBT/Correlations/'
figs_driv <- "/Main_Figures/Figure7/raw_images/Mutation_overlap/withTURBT/Drivers/"
fig_source <- '/Main_Figures/Figure7/source_data/withTURBT/'

supp <- '/Revision_1/Supplementary_Figures/Supplementary_Figure20_cfDNA_old16/mut_overlap_raw_images/withTURBT/'
supp_corr <- '/Revision_1/Supplementary_Figures/Supplementary_Figure20_cfDNA_old16/mut_overlap_raw_images/withTURBT/Correlations/'
supp_source <- '/Revision_1/Supplementary_Figures/Supplementary_Figure20_cfDNA_old16/source_data/'
sup_table <- '/Supplementary_Tables/cfDNA/withTURBT/'
```


annotation/graphing
```{r}
#annotate WES vs WGS
patient <- c('15-109','16-070','16-097','17-026','17-030','17-071','18-101','18-120','19-007','19-044','15-108','16-079','17-020','17-047','18-016','18-065','19-001','19-004','19-022','19-037')
ann_WES <- data.frame(patient = sort(patient), seq_type = c(rep("WES", 7), rep("WGS", 13)))

#colors
mycol <- c("UC"="#815CA6","PUC"="#84C2EB","NE"="#E24D74","UCSD-Focal"="#ACD7A0","UCSD_focal"="#ACD7A0", "UCSD"="#43823D", "UC-sarcomatoid"= "#E7CB94",
           "Founder"= "#FEC216", "Shared"= "#2B3990","Subclonal"="#2B3990", "Private"= "#C91220", 
           "Targeted"="#FD8D3C","Targeted_high"="#FD8D3C", "Targeted_low"="#FD8D3C",  "WGS"="#00BFC4") 
FSP_col <- c("Founder"= "#FEC216", "Shared"= "#3B4F8F", "Private"= "#CE414B") 
pat_col <-c('#e6194b','#3cb44b','#ffe119','#fabed4','#4363d8','#f58231','#42d4f4','#911eb4','#f032e6','#bfef45','#469990','#dcbeff','#9a6324','#fffac8','#800000','#aaffc3','#808000','#ffd8b1','#000075','#a9a9a9') %>%
  setNames(c('15-109','16-070','16-097','17-026','17-030','17-071','18-101','18-120','19-007','19-044','15-108','16-079','17-020','17-047','18-016','18-065','19-001','19-004','19-022','19-037'))

theme_set(theme_bw() + 
            theme(axis.text = element_text(color="black"),
                  rect = element_rect(fill = NULL),
                  axis.ticks = element_line(color="black")))

my_classic <- theme_classic() + 
            theme(axis.text = element_text(color="black"),
                  rect = element_rect(fill = NULL),
                  axis.ticks = element_line(color="black"))
```

collapse to unique mutation per patient
-powered = power > 0.8 from any tumor
-TITAN prioritized power (including TURBT)
-tumor SNVs filtered to pyclone BED file
-annotated: clonality, driver/pathogenic, whether mutation is in TPbed
```{r}
collapsed_power_TITAN <- read.csv("/cfDNA/withTURBT/cfDNAPower_TITANTFx_pycloneBEDfiltered_TumorMuts_withTURBT_Collapsed_TPbed.csv")
```

sample info
```{r}
#tumor sample info
sample_info <- read.csv('/Revision_1/Supplementary_tables/all_sequencing_detailed_March2026.csv')

#TFx
TFx_cfDNA <- collapsed_power_TITAN %>% distinct(cfSample, patient, type, cfDNA_TFx)

#clinical 
clinical <- read.csv('/Supplementary_Tables/clinical_data/all_clinical_data_20260413_forSup.csv')

#cfDNA sample info
cfDNA_sample_info <- read.csv('/Supplementary_Tables/cfDNA/cfDNA_sample_info.csv')
```

### SET UP CAPTURE DFs ###

set up capture dfs
```{r}
WGS_pats <- filter(TFx_cfDNA, type=="WGS")$patient

CAC3_bycfSample <- collapsed_power_TITAN %>% filter(in_TPBED=="Yes") %>%
  filter(cfSample!="19-001_1095_ULP-WGS") %>% #remove 19-001 serum sample, only keep 19-001 plasma sample
  select(cfSample, CAC3) %>% table() %>% as.data.frame() %>%
  pivot_wider(names_from = CAC3, values_from = Freq) %>%
  mutate(total = CAC_called_3 + CAC_under_3, 
         percent = CAC_called_3/(CAC_called_3 + CAC_under_3)) %>%
  left_join(TFx_cfDNA) %>%
  left_join(ann_WES) %>%
  mutate(type = factor(type, levels=c("WGS", "Targeted")), 
         highlow = case_when(patient %in% WGS_pats~"High", .default = "Low"),
         type_highlow = case_when(type=="WGS"~"WGS", type=="Targeted"&highlow=="High"~"Targeted_high", type=="Targeted"&highlow=="Low"~"Targeted_low"),
         cluster_status="Total", 
         type_highlow=factor(type_highlow, levels=c("WGS", "Targeted_high", "Targeted_low")))
#write.csv(CAC3_bycfSample, paste0(sup_table, "CaptureProportion_bySample.csv"), row.names = F)


CAC3_byClonality <- collapsed_power_TITAN %>% filter(in_TPBED=="Yes") %>%
  filter(cfSample!="19-001_1095_ULP-WGS") %>% #remove 19-001 serum sample, only keep 19-001 plasma sample
  select(cfSample,cluster_status, CAC3) %>% table() %>% as.data.frame() %>%
  pivot_wider(names_from = CAC3, values_from = Freq) %>%
  mutate(total = CAC_called_3 + CAC_under_3, 
         percent = CAC_called_3/(CAC_called_3 + CAC_under_3)) %>%
  left_join(TFx_cfDNA) %>%
  left_join(ann_WES) %>%
  mutate(cluster_status = factor(cluster_status, levels=c("Founder", "Shared", "Private")), 
         highlow = case_when(patient %in% WGS_pats~"High", .default = "Low"),
         type_highlow = case_when(type=="WGS"~"WGS", type=="Targeted"&highlow=="High"~"Targeted_high", type=="Targeted"&highlow=="Low"~"Targeted_low"),
         type=factor(type, levels=c("WGS", "Targeted")), 
         type_highlow=factor(type_highlow, levels=c("WGS", "Targeted_high", "Targeted_low")))
#write.csv(CAC3_byClonality, paste0(sup_table, "CaptureProportion_bySample_byFSP.csv"), row.names = F)

##df for all:
all_df <- rbind(CAC3_bycfSample, CAC3_byClonality) %>%
  mutate(cluster_status = factor(cluster_status, levels=c("Total", "Founder", "Shared", "Private")), 
         type=factor(type, levels=c("WGS", "Targeted")))
```

set up CP information
```{r}
CP <- read_csv('/Metastatic_Seeding/patient_cluster_CP_withTURBT.csv', show_col_types = F)
sample_CP <- read_csv('/Metastatic_Seeding/sample_cluster_CP_withTURBT.csv', show_col_types = F)

private_samples <- sample_CP %>% filter(cellular_prevalence>.03, cluster_type=="Private") %>%
  mutate(Block_Letter = case_when(patient %in% c("15-108", "15-109", "16-070", "16-079", "16-097", "17-020", "17-026")~substr(sample_id, 8,8), 
                                  .default = substr(sample_id, 7,7)))

CAC3_byCluster <- collapsed_power_TITAN %>% filter(in_TPBED=="Yes") %>%
  filter(cfSample!="19-001_1095_ULP-WGS") %>%
  select(cfSample, cluster_id, CAC3) %>% table() %>% as.data.frame() %>%
  pivot_wider(names_from = CAC3, values_from = Freq) %>%
  mutate(total = CAC_called_3 + CAC_under_3, 
         percent = CAC_called_3/(CAC_called_3 + CAC_under_3), 
         cluster_id = as.numeric(as.character(cluster_id))) %>%
  left_join(TFx_cfDNA) %>%
  left_join(ann_WES) %>%
  left_join(CP %>% select(patient, cluster_id, cluster_type, cellular_prevalence, cluster_mut_no, n_CP3)) %>%
  filter(!is.na(cluster_type)) %>%
  mutate(weighted_CP = n_CP3*cellular_prevalence, 
         weighted_CP_TFx = n_CP3*cellular_prevalence*cfDNA_TFx, 
         weighted_CP_TFx_num = n_CP3*cellular_prevalence*cfDNA_TFx/cluster_mut_no, 
         cluster_type = factor(cluster_type, levels=c("Founder", "Shared", "Private")), 
         type=factor(type, levels=c("WGS", "Targeted")))
#write.csv(CAC3_byCluster, paste0(sup_table, "CaptureProportion_bySample_byClone.csv"), row.names = F)
```

set up pathology information
```{r message=FALSE, warning=FALSE}
necrosis <- read_csv('/Pathology/Purity_Necrosis.csv', show_col_types = F)[,1:7]
necrosis_allSamples <- read.csv('/Supplementary_Tables/clinical_data/pathology_byBlock.csv')
necrosis_patient <- read.csv('/Supplementary_Tables/clinical_data/pathology_byPatient.csv')

necrosis_plot <- CAC3_byCluster %>%
  left_join(necrosis_patient, by=c("patient"="Patient")) 

#basic info on private mut capture -Table S8 sheet PrivateSNV_byTumor
private_capture <- CAC3_byCluster %>% filter(cellular_prevalence>0.03, cluster_type=="Private") %>%
  left_join(private_samples %>% select(patient, sample_id, cellular_prevalence, cluster_id, cluster_type, Block_Letter)) %>%
  left_join(sample_info %>% select(patient, block_letter, tumor_site, tumor_site_group), by=c("patient"="patient", "Block_Letter"="block_letter")) %>%
  select(patient, sample_id, Block_Letter, tumor_site, tumor_site_group, cfSample, cluster_id, CAC_called_3, CAC_under_3, total, percent, type, cfDNA_TFx, seq_type, cluster_type, cellular_prevalence) %>%
  arrange(patient, sample_id)
write.csv(private_capture, '/Supplementary_Tables/cfDNA/withTURBT/privateCapture_byTumor.csv', row.names = F)

#get necrosis percent specific to each mut cluster (for the tumors that cluster was actually in)
necrosis_plot_private <- necrosis_plot %>% filter(cluster_type=="Private") %>% #only looking at private cluster capture
  left_join(private_samples %>% select(patient, sample_id, cellular_prevalence, cluster_id, cluster_type, Block_Letter)) %>% #pull the sample that private cluster is in
  select(-Percent_Cancer, -Percent_Necrosis) %>%
  left_join(necrosis_allSamples, by=c("patient"="Patient", "Block_Letter"="Block_Letter")) %>% #pull pathology information for that tumor
  left_join(sample_info %>% select(patient, block_letter, tumor_site_group), by=c("patient"="patient", "Block_Letter"="block_letter")) %>%
  mutate(tumor_site_group= case_when(is.na(tumor_site_group)~"Primary", .default = tumor_site_group))

#write.csv(necrosis_plot_private %>% select(patient, cfSample, sample_id, type,CAC_called_3, CAC_under_3, total, percent, Percent_Cancer), paste0(fig_source,'7E_privateCapture_byTumor_pathology.csv'), row.names = F)
```

set up dfs of driver mut capture
```{r}
#deleterious in updated driver gene
CAC3_DelDriver <- collapsed_power_TITAN %>% filter(in_TPBED=="Yes") %>%
  filter(pathogenic_3m_updated=="Deleterious", IsDriverBLCAupdated=="Yes") %>%
  filter(cfSample!="19-001_1095_ULP-WGS") %>% #remove 19-001 serum sample, only keep 19-001 plasma sample
  select(cfSample, CAC3) %>% table() %>% as.data.frame() %>%
  pivot_wider(names_from = CAC3, values_from = Freq) %>%
  mutate(total = CAC_called_3 + CAC_under_3, 
         percent = CAC_called_3/(CAC_called_3 + CAC_under_3)) %>%
  left_join(TFx_cfDNA) %>%
  left_join(ann_WES) %>%
  mutate(type = factor(type, levels=c("WGS", "Targeted")), 
         highlow = case_when(patient %in% WGS_pats~"High", .default = "Low"),
         type_highlow = case_when(type=="WGS"~"WGS", type=="Targeted"&highlow=="High"~"Targeted_high", type=="Targeted"&highlow=="Low"~"Targeted_low"),
         cluster_status="Total", 
         type_highlow=factor(type_highlow, levels=c("WGS", "Targeted_high", "Targeted_low")))
CAC3_DelDriver %>% group_by(type_highlow) %>% summarise(med=median(percent)) #WGS = 92.9%, TP_high = 100%, TP_low = 63.6% -> no change!
CAC3_DelDriver %>% group_by(type_highlow) %>% summarise(total=sum(total), called = sum(CAC_called_3)) #WGS = 35/40 (88%), TP_high = 23/25 (92%), TP_low = 20/32 (63%) -> no change!
#write.csv(CAC3_DelDriver, paste0(sup_table,'CaptureProportion_bySample_DelDriver.csv'), row.names = F)
```

### GRAPHING ### 

calculate medians
```{r}
#total: WGS = 70%, TP = 50% - no change!
CAC3_bycfSample %>%
  group_by(type) %>%
  summarise(med = median(percent))
#total: WGS = 70%, TP_high = 71%, TP_low = 34% - no change!
CAC3_bycfSample %>%
  group_by(type_highlow) %>%
  summarise(med = median(percent))

#by Clonality
## WGS: F = 97.9%, S = 90.1%, P = 28.3%
## TP : F = 81.6%, S = 27.3%, P = 15.3% - changed
CAC3_byClonality %>%
  group_by(type, cluster_status) %>%
  summarise(med = median(percent, na.rm=T))

## WGS.    : F = 97.9%,  S = 90.1%, P = 28.3%
## TP_high : F = 100.0%, S = 78.6%, P = 31.5%
## TP_low  : F = 65.9%,  S = 25.0%, P = 10.0% - changed
CAC3_byClonality %>%
  group_by(type_highlow, cluster_status) %>%
  summarise(med = median(percent, na.rm=T))
```

graph overall CAC3 capture
```{r}
## Fig 7A
CAC3_bycfSample %>% 
  ggplot(aes(x=type_highlow, y=percent, group=type_highlow)) +
    geom_boxplot(aes(fill=type_highlow), show.legend = F, outlier.shape = NA) +
    geom_quasirandom(aes(fill=patient), size=3, height = 0, width=0.25, shape=21) +
    scale_fill_manual(values=c(mycol, pat_col), breaks=names(pat_col)) +
    guides(fill=guide_legend(ncol=2, override.aes = list(shape = 21))) +
    scale_y_continuous(limits = c(0,1), breaks=seq(0,1,0.2), minor_breaks = NULL) +
   labs(title="% of all mutations captured by cfDNA", x="", y="Capture Proportion", fill="Patient", shape="Tumor Seq Type")
ggsave(paste0(wd, "CAC3_all_TPBED150filtered_byType_highlowTFxBox.pdf"), width = 5, height=4)
ggsave(paste0(figs, "CAC3_all_TPBED150filtered_byType_highlowTFxBox.pdf"), width = 4.5, height=4)
```

graph CAC3 capture by clonality
```{r}
#separate high low TFx Fig 7B
stat.test <- CAC3_byClonality %>%
  group_by(type_highlow) %>%
  wilcox_test(percent~cluster_status) %>%
  add_xy_position(step.increase = 0.1)
CAC3_byClonality %>%
  ggplot(aes(x=cluster_status, y=percent, group=cluster_status)) +
    geom_boxplot(aes(fill=cluster_status), show.legend = F, outlier.shape = NA, color="black") +
    stat_summary(fun = median, geom = "crossbar", width = 0.7, color = "white") +
    geom_quasirandom(aes(fill=patient), size=3, shape=21) +
    scale_fill_manual(values=c(mycol, pat_col), breaks=names(pat_col)) +
    guides(fill=guide_legend(ncol=2, override.aes = list(shape = 21))) +
    #stat_pvalue_manual(stat.test, tip.length = 0, label="p.adj") +
    facet_grid(cols=vars(type_highlow)) +
    scale_y_continuous(limits = c(0,1), breaks=seq(0,1,0.2), minor_breaks = NULL) +
   labs(title="% of all mutations captured by cfDNA", x="", y="Capture Proportion", fill="Patient", shape="Tumor Seq Type")
ggsave(paste0(wd, "CAC3_all_TPBED150filtered_byType_byClonality_highlowTFxBox.pdf"), width = 7, height=4)
ggsave(paste0(figs, "CAC3_all_TPBED150filtered_byType_byClonality_highlowTFxBox_withstats.pdf"), width = 7, height=3.5)
```

### CORRELATIONS ###

graph capture-TFx correlation
```{r}
#separate TP vs WGS by clonality **S20G**
CAC3_byClonality %>%
  ggplot(aes(x=cfDNA_TFx, y=percent)) +
    geom_smooth(method=lm, color="black", fill="snow2") +
    geom_point(aes(fill=patient), size=3, shape=21) +
    #stat_cor(label.x.npc = 0, label.y = 0.1,method = "spearman") +
    scale_fill_manual(values=c(mycol, pat_col)) +
    guides(fill=guide_legend(ncol=2, override.aes = list(shape = 21))) +
    scale_y_continuous(breaks=seq(0,1,0.2), minor_breaks = NULL) +
    coord_cartesian(ylim=c(0,1)) +
    facet_grid(rows=vars(type), cols = vars(cluster_status)) +
    labs(title="% of all mutations captured by cfDNA", x="cfDNA TFx", y="Capture Proportion", fill ="Patient", caption="spearman corr")
ggsave(paste0(wd_corr, "CAC3_all_TPBED150filtered_correlation_TITANTFx_byType_byClonality.pdf"), width = 8, height=6)
ggsave(paste0(supp_corr, "CAC3_TITANTFx_byType_byClonality_nostats.pdf"), width = 8, height=4.5)
```

CP correlation
```{r}
#correlation with CP **S20H**
CAC3_byCluster %>%
  ggplot(aes(x=cellular_prevalence, y=percent)) +
    geom_smooth(method=lm, color="black", fill="snow2") +
    geom_point(aes(fill=patient), size=3, shape=21) +
    #stat_cor(label.x.npc = 0, label.y = -0.1, method = "spearman") +
    scale_fill_manual(values=c(mycol, pat_col)) +
    guides(fill=guide_legend(ncol=2, override.aes = list(shape = 21))) +
    coord_cartesian(ylim=c(-0.15,1.05), xlim=c(-0.05,1.05)) + 
    facet_grid(rows=vars(type), cols = vars(cluster_type)) +
    labs(x="Clone CP", title="Capture Correlation to CP per clone")
ggsave(paste0(wd_corr, "CAC3_byClone_byType_byClonality_correlationCP.pdf"), width = 8, height=5)
ggsave(paste0(supp_corr, "CAC3_byClone_byType_byClonality_correlationCP_nostats.pdf"), width = 8, height=4.5)

#correlation with clone abundance*TFx (CP * # tumors * TFx)
## FSP 
CAC3_byCluster %>%
  ggplot(aes(x=weighted_CP_TFx, y=percent)) +
    geom_smooth(method=lm, color="black", fill="snow2") +
    geom_point(aes(fill=patient), size=3, shape=21) +
    stat_cor(label.x.npc = 0, label.y = -0.1, method = "spearman") +
    scale_fill_manual(values=c(mycol, pat_col)) +
    guides(fill=guide_legend(ncol=2, override.aes = list(shape = 21))) +
    coord_cartesian(ylim=c(-0.15,1.05)) +
    facet_grid(rows=vars(type), cols = vars(cluster_type), scales = "free_x") +
    labs(x="weighted_CP_TFx (CP * # tumors * TFx)")
ggsave(paste0(wd_corr, "CAC3_byClone_byType_byClonality_correlationWeightedCPTITANTFx.pdf"), width = 8, height=5)
ggsave(paste0(fig_corr, "CAC3_byClone_byType_byClonality_correlationWeightedCPTITANTFx.pdf"), width = 8, height=5)

## JUST SHARED **7D**
CAC3_byCluster %>%
  filter(cluster_type =="Shared", percent!="NaN") %>% 
  mutate(type = factor(type, levels=c("WGS", "Targeted"))) %>%
  ggplot(aes(x=weighted_CP_TFx, y=percent)) +
    geom_smooth(method=lm, color="black", fill="snow2") +
    geom_point(aes(fill=patient), size=3, shape=21) +
    #stat_cor(label.x.npc = 0, label.y = -0.1, method = "spearman") + coord_cartesian(ylim=c(-0.15,1.05)) +
    scale_y_continuous(breaks=seq(0,1,0.2)) + coord_cartesian(ylim=c(0,1)) +
    scale_fill_manual(values=c(mycol, pat_col)) +
    guides(fill=guide_legend(ncol=2, override.aes = list(shape = 21))) +
    facet_grid(cols=vars(type), scales = "free_x") +
    labs(x="weighted_CP_TFx (CP * # tumors * TFx)")
ggsave(paste0(wd_corr, "CAC3_byClone_byType_SHARED_correlationWeightedCPTITANTFx.pdf"), width = 6, height=2.5)
ggsave(paste0(fig_corr, "CAC3_byClone_byType_SHARED_correlationWeightedCPTITANTFx_nostats.pdf"), width = 6, height=2.5)
```

time death to autopsy
```{r}
df <- CAC3_bycfSample %>% 
  left_join(clinical %>% select(Patient, pass_to_TAN_start_hr), by=c("patient"="Patient")) %>% 
  left_join(cfDNA_sample_info %>% select(Patient, Yield..ng.mL.), by=c("patient"="Patient")) %>%
  select(cfSample, patient, type, percent, pass_to_TAN_start_hr, cfDNA_TFx, Yield..ng.mL.)
write.csv(df, paste0(supp_source, "20BCD_Sample_Capture_TimeToTAN.csv"), row.names = F)

#TFx (per patient) ** S20B **
df %>%
  distinct(patient, pass_to_TAN_start_hr, cfDNA_TFx) %>%
  ggplot(aes(x=pass_to_TAN_start_hr, y=cfDNA_TFx)) +
    geom_smooth(method=lm, color="black", fill="snow2") +
    geom_point(size=3) +
    stat_cor(label.x.npc = 0, label.y = -0.1, method = "spearman") 

#yield (per patient) ** S20C **
df %>%
  distinct(patient, pass_to_TAN_start_hr, Yield..ng.mL.) %>%
  ggplot(aes(x=pass_to_TAN_start_hr, y=Yield..ng.mL.)) +
    geom_smooth(method=lm, color="black", fill="snow2") +
    geom_point(size=3) +
    stat_cor(label.x.npc = 0.5, label.y.npc = 0.8, method = "spearman") +
    coord_cartesian(ylim=c(0,6600)) +
    scale_y_continuous(breaks=seq(0,6000,2000), labels = seq(0,6,2), name = "Yield ug/mL")

#capture (per cfDNA sample) ** S20D **
df %>%
  ggplot(aes(x=pass_to_TAN_start_hr, y=percent)) +
    geom_smooth(method=lm, color="black", fill="snow2") +
    geom_point(aes(color=type), size=3) +
    stat_cor(label.x.npc = 0, label.y = -0.1, method = "spearman") +
    scale_color_manual(values=mycol)
```


### VOLUME + PATHOLOGY ###

CT tumor volume
```{r}
#find tumor volumes for every tumor sequenced at death that has a private clone from latest CT scan within 3.5 months of death - somewhat manually curated
CT_vol <- read.csv('/CT Imaging/CT_tumor_volumes_cfDNA_withTURBT.csv')

#combine volumes with private TP capture + pathology 
CT_vol_all <- CT_vol %>% 
  left_join(necrosis_plot_private %>% filter(type=="Targeted")) %>%
  filter(!is.na(Volume_cc), !is.nan(percent))

CT_vol_collapsed <- CT_vol_all %>%
  group_by(patient, Volume_cc) %>% 
  summarise(percent = mean(percent), 
            cfDNA_TFx = unique(cfDNA_TFx), 
            cellular_prevalence = mean(cellular_prevalence), 
            #weighted_CP = mean(weighted_CP), 
            #weighted_CP_TFx = mean(weighted_CP_TFx), 
            Percent_Cancer = mean(Percent_Cancer), 
            #Percent_Necrosis = mean(Percent_Necrosis), 
            tumor_site_group = unique(tumor_site_group), .groups = "drop")
write.csv(CT_vol_collapsed, paste0(fig_source, "7F_CTvol_privateCapture.csv"), row.names = F)
```

correlation with CT vol 
```{r}
## per tumor volume / private capture only

## just volume **S20J**
CT_vol_collapsed %>%
  #filter(!(patient =="19-037"&tumor_site_group=="LN")) %>%
  ggplot(aes(x=Volume_cc, y=percent)) +
    geom_smooth(method=lm, color="black", fill="snow2") +
    geom_point(aes(fill=patient), shape=21, size=3) +
    #stat_cor(label.x.npc = 0, label.y.npc = 0.9,method = "spearman") +
    scale_fill_manual(values=c(mycol, pat_col)) +
    scale_shape_manual(values=c(21:25)) +
    guides(fill=guide_legend(ncol=1, override.aes = list(shape = 21))) +
    scale_y_continuous(breaks=seq(0,1,0.2), minor_breaks = NULL) +
    #scale_x_continuous(transform = "log2", breaks=c(2,5,10,50,250,500, 1000), labels = c(2,5,10,50,250,500, 1000), limits = c(1.7,1020)) +
    coord_cartesian(ylim=c(0,1)) +
    labs(y="Capture Proportion", fill ="Patient", caption="spearman corr, n=19 tumors/12pats")
ggsave(paste0(wd_corr, "CAC3_CTvol_Private_correlation_Volume.pdf"), width=4.25, height=3)
ggsave(paste0(supp_corr, "CAC3_CTvol_Private_correlation_Volume_nostats.pdf"), width=3.6, height=3)


#scan volume * pathology purity **7F**
CT_vol_collapsed %>%
  #filter(!(patient =="19-037"&tumor_site_group=="LN")) %>% 
  ggplot(aes(x=Volume_cc*Percent_Cancer/100, y=percent)) +
    geom_smooth(method=lm, color="black", fill="snow2") +
    geom_point(aes(fill=patient), size=4, shape=21, alpha=0.8) +
    stat_cor(label.x.npc = 0, label.y.npc = 0.9,method = "spearman") +
    scale_fill_manual(values=c(mycol, pat_col)) +
    guides(fill=guide_legend(ncol=1, override.aes = list(shape = 21))) +
    scale_y_continuous(breaks=seq(0,1,0.2), minor_breaks = NULL) +
    #scale_x_continuous(transform = "log2", breaks=c(0.05,0.5,1,2,5,10,50,250,500, 1000), labels = c(0.05,0.5,1,2,5,10,50,250,500, 1000), limits = c(0.05,700)) +
    coord_cartesian(ylim=c(0,1)) +
    labs(y="Capture Proportion", fill ="Patient", caption="spearman corr, n=19 tumors/12pats")
ggsave(paste0(wd_corr, "CAC3_CTvol_Private_correlation_VolumePurity.pdf"), width=6, height=4)
ggsave(paste0(fig_corr, "CAC3_CTvol_Private_correlation_VolumePurity_nostats.pdf"), width=3.6, height=3)
```

pathology - tumor specific (private only)
```{r}
# % cancer **7E**
necrosis_plot_private %>%
  mutate(type = factor(type, levels=c("WGS", "Targeted"))) %>% 
  filter(percent!="NaN") %>% 
  ggplot(aes(x=Percent_Cancer, y=percent)) +
    geom_smooth(method=lm, color="black", fill="snow2") +
    geom_point(aes(fill=patient), size=3, shape=21) +
    guides(fill=guide_legend(ncol=2, override.aes = list(shape = 21))) +
    #stat_cor(label.x.npc = 0, label.y = -0.1, method = "spearman") + coord_cartesian(ylim=c(-0.15,1.05)) +
    coord_cartesian(ylim=c(0,1)) +
    scale_fill_manual(values=c(mycol, pat_col)) +
    facet_grid(cols=vars(type), scales = "free_x") +
    scale_y_continuous(breaks=seq(0,1,0.2)) + 
    scale_x_continuous(breaks=seq(0,100,20)) +
    labs(y="Private capture proportion")
ggsave(paste0(wd_corr, "CAC3_byClone_Privates_correlationPathPurity.pdf"), width = 6, height=2.5)
ggsave(paste0(fig_corr, "CAC3_byClone_Privates_correlationPathPurity_nostats.pdf"), width = 6, height=2.5)
```

tumor site
```{r}
df <- necrosis_plot_private %>% 
  filter(percent!="NaN") %>%
  mutate(type = factor(type, levels=c("WGS", "Targeted")), 
         tumor_site_group = factor(tumor_site_group, levels=c("Primary", "Liver", "LN", "Lung", "Other")), 
         highlow = case_when(patient %in% WGS_pats~"High", .default = "Low"),
         type_highlow = case_when(type=="WGS"~"WGS", type=="Targeted"&highlow=="High"~"Targeted_high", type=="Targeted"&highlow=="Low"~"Targeted_low"), 
         type_highlow = factor(type_highlow, levels=c("WGS", "Targeted_high", "Targeted_low")))

write.csv(df, paste0(supp_source,'S20I_private_byTumorSite.csv'), row.names = F)
 
#separate WGS, TP **S20I**
stat.test <- df %>%
  group_by(type) %>%
  wilcox_test(percent~tumor_site_group) %>%
  add_xy_position(step.increase = 0.02)

df %>%
  ggplot(aes(x=tumor_site_group, y=percent)) +
    geom_boxplot(fill="snow1", outlier.shape = NA) +
    geom_jitter(width=0.2, height=0, size=3, shape=21, aes(fill=patient), show.legend = F) +
    facet_wrap(~type) +
    my_classic + border() +
    #stat_pvalue_manual(stat.test %>% filter(p<0.1), label="p.adj", tip.length = 0) + 
    scale_fill_manual(values=pat_col) +
    scale_y_continuous(breaks=seq(0,1,0.2)) +
    labs(y="Private Capture Proportion", caption="no sig by wilcox")
ggsave(paste0(wd, "PrivateCapture_byTumorSite.pdf"), width=6, height=3.5)
ggsave(paste0(supp, "PrivateCapture_byTumorSite_nostats.pdf"), width=6, height=3.5)
```


### DRIVERS ###

graphing driver capture 
```{r}
#deleterious mut in driver gene
CAC3_DelDriver %>%
  ggplot(aes(x=type_highlow, y=percent, group=type_highlow)) +
    geom_boxplot(aes(fill=type_highlow), show.legend = F,width=0.6, outlier.shape = NA, color="black") +
    stat_summary(fun = median, geom = "crossbar", width = 0.6, color = "white") +
    geom_quasirandom(aes(fill=patient), size=3, shape=21) +
    scale_fill_manual(values=c(mycol, pat_col), breaks=names(pat_col)) +
    guides(fill=guide_legend(ncol=2, override.aes = list(shape = 21))) +
    scale_y_continuous(limits = c(0,1), breaks=seq(0,1,0.2), minor_breaks = NULL) +
   labs(title="Deleterious BLCA Driver SNVs Capture", x="", y="Capture Proportion", fill="Patient")
ggsave(paste0(wd_driv, "DelDriver_CAC3_byHighLowType_noX_Box.pdf"), width=5, height=4)
ggsave(paste0(figs_driv, "DelDriver_CAC3_byHighLowType_noX_Box.pdf"), width=5, height=4)
```

comut for deleterious drivers in InterestingBLCA genes (w/ TPbed annotated)
```{r}
#having TP & WGS together on same comut
del_driver2 <- collapsed_power_TITAN %>% 
  filter(IsDriverBLCAupdated=="Yes", pathogenic_3m_updated=="Deleterious", cfSample !="19-001_1095_ULP-WGS") %>%
  mutate(type = ifelse(grepl("cfWGS", cfSample), "WGS", "Targeted"), 
         patient = substr(cfSample, 1, 6)) 
GOI <- del_driver2 %>% pull(Gene.refGene) %>% unique()
cfSamples <- unique(del_driver2$cfSample)

NOI <- del_driver2 %>% 
  group_by(Gene.refGene) %>%
  summarise(n=n()) %>%
  filter(n==1) %>% pull(Gene.refGene)
NOI <- c(NOI, "MACF1", "ZFP36L1", "TAF15", "RXRA", "GNA13", "RXRA", "TP63", "PREX2", "MYH9", "HSPG2", "CHD2", "BIRC6", "ARHGEF10")
GOI <- setdiff(GOI, NOI)

comut <- data.frame()
for(gene in GOI){
  for(samp in cfSamples){
    in_tumor <- ifelse(del_driver2 %>% filter(Gene.refGene==gene, cfSample==samp) %>% nrow() == 0, "no_tumor_mut", "tumor_mut")
    inTPbed <- ifelse(in_tumor=="tumor_mut", 
                      del_driver2 %>% filter(Gene.refGene==gene, cfSample==samp) %>% pull(in_TPBED), "no_tumor_mut")
    clonality <- ifelse(in_tumor=="tumor_mut", 
                        del_driver2 %>% filter(Gene.refGene==gene, cfSample==samp) %>% pull(cluster_status), "no_tumor_mut")
    in_cfDNA <- ifelse(del_driver2 %>% filter(Gene.refGene==gene, cfSample==samp, CAC3=="CAC_called_3") %>% nrow() == 0, 
                       "not_in_cfDNA", 
                       del_driver2 %>% filter(Gene.refGene==gene, cfSample==samp, CAC3=="CAC_called_3") %>% pull(cluster_status))
    patient <- substr(samp, 1, 6)
    type = ifelse(grepl("cfWGS", samp), "WGS", "Targeted")
    df <- data.frame(Gene = gene, cfSample = samp,patient = patient,type=type, in_tumor = clonality, in_cfDNA = in_cfDNA, in_TPBED = inTPbed)
    
    comut <- bind_rows(comut, df)
  }
}
#write.csv(comut, '/Supplementary_Tables/cfDNA/cfDNA_driver_comut.csv', row.names = F)

order_genes <- comut %>% filter(in_tumor!="no_tumor_mut", in_TPBED=="Yes") %>% pull(Gene) %>% table() %>% as.data.frame() %>% arrange(desc(Freq)) %>% rename("Gene"=".") %>% mutate(Gene = as.character(Gene)) %>% pull(Gene)
order_genes <- c(order_genes, setdiff(unique(comut$Gene), order_genes))
order_samples_WGS <- comut %>% filter(in_tumor!="no_tumor_mut", type=="WGS") %>% pull(cfSample) %>% table() %>% as.data.frame() %>% arrange(desc(Freq)) %>% rename("cfSample"=".") %>% mutate(cfSample = as.character(cfSample))%>% pull(cfSample)

order_patient <- substr(order_samples_WGS, 1, 6)
order_patient <- c("19-022", "15-109", "19-037", "16-070", "16-079", "16-097", "18-101", "19-001")

order_samples <- comut %>% filter(patient %in% order_patient) %>% distinct(patient, type, cfSample) %>% mutate(patient = factor(patient, levels=order_patient)) %>% arrange(patient) %>% pull(cfSample)
other_TP_samples <- comut %>% filter(type=="Targeted", !patient %in% order_patient, in_tumor!="no_tumor_mut") %>% pull(cfSample) %>% table() %>% as.data.frame() %>% arrange(desc(Freq)) %>% rename("cfSample"=".") %>% mutate(cfSample = as.character(cfSample))%>% pull(cfSample)
order_samples <- append(order_samples, other_TP_samples)

a <- length(GOI) +1
b <- comut %>%
  mutate(Gene = factor(Gene, levels=rev(order_genes)), 
         cfSample = factor(cfSample, levels=order_samples)) %>% filter(Gene=="BRCA2") %>% arrange(cfSample) %>% pull(patient)

comut_order <- comut %>%
  mutate(Gene = factor(Gene, levels=rev(order_genes)), 
         cfSample = factor(cfSample, levels=order_samples))


ggplot() +
  
  geom_tile(data = comut_order %>% filter(in_TPBED!="No"), aes(x=cfSample, y=Gene, color=in_tumor, fill=in_cfDNA),width=0.5, height=0.5, linewidth =1.5) +
  geom_tile(data = comut_order %>% filter(in_TPBED=="No"), aes(x=cfSample, y=Gene), color="snow2", fill="white", width=0.4, height=0.4,linewidth=1) +
  
  geom_vline(xintercept = seq(2,16,2)+0.5, color="snow3", linetype="dashed") +
  scale_color_manual(values=c("no_tumor_mut"="snow2", mycol), breaks = c("Founder", "Shared","Private", "no_tumor_mut"), labels=c("Founder", "Shared","Private", "No tumor mut")) +
  scale_fill_manual(values=c("not_in_cfDNA"="snow2", mycol), breaks = c("Founder", "Shared","Private", "not_in_cfDNA"), labels=c("Founder", "Shared","Private", "Not in cfDNA")) +
  
  my_classic +
  theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust=1)) +
  scale_x_discrete(labels = b) +
  geom_tile(data = comut_order, aes(x=cfSample, fill=type, color=type, y=a), width=0.5, height=0.5, inherit.aes = F, linewidth = 1.5) +
  labs(title="Deleterious BLCA Driver Mutation Capture in cfWGS", fill="In cfDNA", color="Tumor Clonality", x="Patient")
ggsave(paste0(wd_driv, "Comut_DelDriver_over1.pdf"), width=7, height=6)
ggsave(paste0(figs_driv, "Comut_DelDriver_over1.pdf"), width=7, height=6)
```


### CAPTURE EACH TUMOR ### 

import/set up allpower df
-use allpower, since we're looking at private mutations in each tumor, we don't want to use the collapsed mutations, but retain the call for each tumor
```{r}
#original: per mutation, per tumor power calculation
allpower <- read.csv('/cfDNA/withTURBT/cfDNAPower_TITANTFx_pycloneBEDfiltered_TumorMuts_withTURBT.csv')

allpower <- allpower %>%
  left_join(collapsed_power_TITAN %>% distinct(id, cfSample, in_TPBED))
```

are there private mutations to call in each tumor?
```{r}
#all private mutations - capture in cfDNA
df <- allpower %>% filter(cluster_status=="Private") %>%
  filter(cfSample!="19-001_1095_ULP-WGS") %>% #remove 19-001 serum sample, only keep 19-001 plasma sample
  select(cfSample, sample, CAC3) %>% table() %>% as.data.frame() %>%
  pivot_wider(names_from = CAC3, values_from = Freq) %>%
  mutate(hasPrivate = ifelse(CAC_called_3>0, "Yes", "No"), 
         type = ifelse(grepl("cfWGS", cfSample), "WGS", "Targeted"), 
         type = factor(type, levels=c("WGS", "Targeted")),
         patient = substr(sample, 1, 6)) %>%
  rowwise %>% filter(grepl(patient, cfSample)) %>% ungroup() %>%
  left_join(ann_WES)

#subset TP to only mutations in TP region
df2 <- allpower %>% filter(cluster_status=="Private", in_TPBED =="Yes") %>%
  filter(cfSample!="19-001_1095_ULP-WGS") %>% #remove 19-001 serum sample, only keep 19-001 plasma sample
  select(cfSample, sample, CAC3) %>% table() %>% as.data.frame() %>%
  pivot_wider(names_from = CAC3, values_from = Freq) %>%
  mutate(hasPrivate = ifelse(CAC_called_3>0, "Yes", "No"), 
         type = ifelse(grepl("cfWGS", cfSample), "WGS", "Targeted"), 
         type = factor(type, levels=c("WGS", "Targeted")),
         patient = substr(sample, 1, 6)) %>%
  rowwise %>% filter(grepl(patient, cfSample)) %>% ungroup() %>%
  left_join(ann_WES)

#for cfWGS, 35/35 tumors are found
df %>% filter(type=="WGS")

#for cfTP, there are 70 tumors possible -> they all have some mutations, but only 39 have mutations in TP region. 17/39 are captured in cfDNA
df %>% filter(type =="Targeted") %>%
  mutate(total_mut = CAC_called_3 + CAC_under_3) %>%
  arrange(total_mut) 
df2 %>% filter(type =="Targeted") %>%
  mutate(total_mut = CAC_called_3 + CAC_under_3) %>%
  arrange(total_mut) %>% 
  filter(total_mut>0) %>%
  filter(hasPrivate=="Yes")
```


per tumor basis
```{r}
my_classic <- theme_classic() + 
            theme(axis.text = element_text(color="black"),
                  rect = element_rect(fill = NULL),
                  axis.ticks = element_line(color="black"))

#per tumor sample, are private mutations called?
callEach <- allpower %>% filter(cluster_status=="Private", in_TPBED=="Yes") %>%
  filter(cfSample!="19-001_1095_ULP-WGS") %>% #remove 19-001 serum sample, only keep 19-001 plasma sample
  select(cfSample, sample, CAC3) %>% table() %>% as.data.frame() %>%
  pivot_wider(names_from = CAC3, values_from = Freq) %>%
  mutate(hasPrivate = ifelse(CAC_called_3>0, "Yes", "No"), 
         no_Private = case_when(CAC_called_3<6~as.factor(CAC_called_3), 
                                between(CAC_called_3, 6, 10)~"6-10", 
                                between(CAC_called_3, 11, 20)~"11-20",
                                between(CAC_called_3, 21, 30)~"21-30",
                                between(CAC_called_3, 31, 50)~"31-50",
                                between(CAC_called_3, 51, 100)~"51-100",
                                between(CAC_called_3, 101, 200)~"101-200",
                                between(CAC_called_3,201 , 500)~"201-500",
                                CAC_called_3>500~"501+"),
         type = ifelse(grepl("cfWGS", cfSample), "WGS", "Targeted"), 
         patient = substr(sample, 1, 6)) %>%
  rowwise %>% filter(grepl(patient, cfSample)) %>% ungroup() %>%
  left_join(ann_WES) %>%
  mutate(no_Private = factor(no_Private, levels=c(as.character(0:5), "6-10", "11-20", "21-30", "31-50", "51-100", "101-200", "201-500", "501+")), 
         type = factor(type, levels=c("WGS", "Targeted"))) %>% 
  mutate(total_mut = CAC_called_3 + CAC_under_3) %>%
  filter(total_mut>0)
write.csv(callEach, paste0(fig_source, '7C_callEachTumor.csv'), row.names = F)

callEach %>%
  ggplot(aes(x=type, fill=no_Private)) +
    geom_bar(stat="count", position="fill") +
    scale_fill_manual(values=c("snow2",magma(14)[3:14])) +
    my_classic +
    labs(title="Capture Per Tumor by Private Mutations", x="", y="% of tumors captured", fill="# mutations", subtitle = "captured: 1+ private mutation CAC3")
ggsave(paste0(wd, "PerTumorCapture_Detailed_fill.pdf"), width=3, height=4)
ggsave(paste0(figs, "PerTumorCapture_Detailed_fill.pdf"), width=3, height=4)
```



false positive for per tumor basis
```{r}
#calculate # of false positive private mutations expected per tumor from ctDNA depth & sequencing error rate
#make df with min # of mutations needed per sample to "capture" > false positives
min_capture <- allpower %>% filter(cluster_status=="Private") %>%
  mutate(false_pos = 1- (dbinom(0,Total_reads_ctDNA, 0.001)+dbinom(1,Total_reads_ctDNA, 0.001)+dbinom(2,Total_reads_ctDNA, 0.001))) %>%
  group_by(sample, cfSample) %>%
  summarise(sum_false_pos = sum(false_pos), 
            n_muts = n()) %>% 
  mutate(min_for_capture_001 = ceiling(sum_false_pos)) 


#scatterplot of expected false pos vs CAC3
## error = 0.001
obs_exp <- allpower %>% filter(cluster_status=="Private") %>%
  select(cfSample, sample, CAC3) %>% table() %>% as.data.frame() %>%
  pivot_wider(names_from = CAC3, values_from = Freq) %>%
  mutate(type = ifelse(grepl("cfWGS", cfSample), "WGS", "Targeted"), 
         patient = substr(sample, 1, 6)) %>%
  rowwise %>% filter(grepl(patient, cfSample)) %>% ungroup() %>%
  left_join(min_capture) %>%
  left_join(ann_WES) %>%
  mutate(above = ifelse(CAC_called_3>sum_false_pos, "yes", "no"))
write.csv(obs_exp, paste0(supp_source, 'S20F_Observed_Expected_PrivateCapture.csv'), row.names = F)

obs_exp %>% ## **S20F**
  ggplot(aes(x=sum_false_pos, y=CAC_called_3+0.1)) +
    geom_jitter(aes(fill=above, shape=type), size=4, alpha=0.8, width=0.1, height=0.1) +
    scale_shape_manual(values=c(21,24)) +
    scale_fill_manual(values=c(pat_col, "yes"="forestgreen", "no"="snow3")) +
    guides(fill=guide_legend(ncol=1, override.aes = list(shape = 21))) +
    my_classic + border() +
    scale_x_continuous(breaks=seq(0,10,2)) +
    scale_y_continuous(transform = "log10", breaks=c(0.1,1,10,100,1000), labels = c(0,1,10,100,1000)) +
    geom_line(data = data.frame(x = seq(0.1, 10, length.out = 100)) |> transform(y = x),
            aes(x = x, y = y), color = "black", linetype = "dashed") +
    labs(x="Expected False Positives", y=" Called Mutations", fill="Observed > Expected", shape="cfDNA Seq Type", title="Capture over Expected False Positives", subtitle = "using error = 0.001")
ggsave(paste0(wd, "PerTumor_FalsePositivesvsCAC3_001.pdf"), width=5, height=4)
ggsave(paste0(supp, "PerTumor_FalsePositivesvsCAC3_001.pdf"), width=5.5, height=4)
```

