---
title: "CollectAllelicCounts"
output: html_document
date: "2024-10-29"
---

libraries
```{r message=FALSE, warning=FALSE, include=FALSE}
library(data.table)
library(readxl)
library(rstatix)
library(ggpubr)
library(GenomicRanges)
library(rtracklayer)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(tidyverse)
```

### INPUTS ###

sample df
```{r}
inputs <- read_csv("/cfDNA/With_TopOff/PowerCalc/CAC_sh_inputs.csv") %>%
  select(patient, seqType, ID)

tumor_filtered <- read.csv('/cfDNA/withTURBT/WGS_tissue_SNVs_withTURBT_pycloneBEDfiltered.csv')
```

set up function to read output of CollectAllelicCounts
```{r}
# Function to read output file of CollectAllelicCounts and ignore rows starting with '@', then read the rest into a data frame
readFileIgnoreAtRows <- function(filePath) {
  # Step 1: Read the entire file into a character vector
  fileContent <- readLines(filePath)
  
  # Initialize an empty vector to hold the non-@ lines
  nonAtLines <- c()
  
  # Step 2: Filter out lines starting with '@'
  for(line in fileContent) {
    if (!grepl("^@", line)) { # Using grepl to check if the line does NOT start with '@'
      nonAtLines <- c(nonAtLines, line)
    }
  }
  # Convert the list of lines into a single string, ensuring tabs are preserved
  dataString <- paste(nonAtLines, collapse = "\n")
  
  # Step 3: Convert the string into a data frame, handling tab separation
  # Splitting the string into lines based on newline characters
  dataLines <- strsplit(dataString, "\n")[[1]]
  
  # Converting the list of trimmed lines into a matrix, using tab as separator
  dataMatrix <- do.call(rbind, lapply(dataLines, function(line) strsplit(line, "\\t")[[1]]))
  
  # Converting the matrix to a data frame
  dataFrame <- as.data.frame(dataMatrix, stringsAsFactors = FALSE)
  colnames(dataFrame)=dataFrame[1,]
  dataFrame=dataFrame[-1,]
  return(dataFrame)
}
```

get cfDNA TFx - use TITAN TFx if patient had WGS, ichor TFx for patients without deep WGS
```{r}
#Oct 2024 updated TITAN TFx with Top Off for WGS cfDNA
DEEP_TFx <- read_delim("BC-TAN_TFX_Subtype_cfDNA.tsv", delim = "\t", show_col_types = F) %>%
  mutate(patient = substr(sample, 8,13)) %>%
  left_join(inputs %>% filter(seqType=="cfWGS")) %>%
  rename(sample_id=ID, type=seqType, cfDNA_TFx = TITAN_tfx_topOffs, preTopOff_TFx=tfx, patient_id=patient) %>% 
  mutate(cfSample = paste0(sample_id, "_", type))

#ichor TFx for targeted cfDNA (not updated, but correct- didn't need to rerun ULP w/ July updates)
ULP_TFx <- read_csv("/important_files_References/sample_details_tumorfractions_clinical.csv")[,1:6] %>%
  filter(grepl("ULP", type)) %>%
  rename(cfDNA_TFx = ichor_tfx) %>%
  mutate(cfSample = paste0(sample_id, "_", type))

#combine TFx:
##using TITAN TFx for all if patient had WGS, or ichor TFx for patients without deep WGS
TFx_cfDNA <- data.frame()
samps <- unique(c(DEEP_TFx$cfSample[1:8], ULP_TFx$cfSample))
for(samp in samps){
  pat <- substr(samp, 1, 6)
  has_WGS <- nrow(DEEP_TFx %>% filter(patient_id==pat))>0
  if(has_WGS){TFx = DEEP_TFx %>% filter(patient_id==pat) %>% pull(cfDNA_TFx)} else {TFx = ULP_TFx %>% filter(cfSample==samp) %>% pull(cfDNA_TFx)}
  sample_id <- substr(samp, 1, 11)
  type <- ifelse(grepl("cfWGS", samp), "cfWGS", "ULP-WGS")
  
  temp <- data.frame(patient_id = pat, cfSample = samp, sample_id = sample_id, cfDNA_TFx = TFx, type = type)
  TFx_cfDNA <- bind_rows(TFx_cfDNA, temp)
}

rm(DEEP_TFx, ULP_TFx, temp)
```

get tissue tumor purity
```{r include=FALSE}
#read in - Pushpa made this file
tumor_purity <- read.csv("WGS_purity_ploidy_curated_Feb2026.csv") %>%
  mutate(sample_id= case_match(sample_id, "16-079_Primary-FFPE"~"16-079_Primary-FF", .default = sample_id)) #allpower CAC inputs has "FF" so make tumor_purity match that
```

### COMBINE + POWER CALC ###

do power calc (original, per tumor)
```{r}
#loop over all
mydir <- "/cell_free_dna/OverlapPowerCalc/NEWTopOff/output/"
myfl <- list.files(mydir)
#fl <- myfl[3]

allpower <- data.frame()
for(fl in myfl){
  pat <- substr(fl,1,6)
  #cfDNA read counts in each cfDNA sample for all tumor mutations in that patient
  GAC <- readFileIgnoreAtRows(paste0(mydir,fl)) %>% 
    mutate(across(2:4, as.numeric)) %>%
    `colnames<-`(c("Chr", "Start", "Ref_reads_ctDNA", "Alt_reads_ctDNA", "Ref", "Alt")) %>%
    mutate("Total_reads_ctDNA" = Ref_reads_ctDNA + Alt_reads_ctDNA, 
          varID = paste(Chr, Start, Ref, sep = "_"), 
          cfSample = substr(fl, 1,11))
  
  #add cfDNA TFx to above df, using titan or ichor depending on cfDNA sample type
  TFx <- TFx_cfDNA %>% filter(patient_id == pat)
  if(substr(fl, 13,16)=="cfTP"){
    GAC <- GAC %>% left_join(TFx %>% filter(grepl("ULP-WGS", cfSample)), by=c("cfSample"= "sample_id"))
  } else{GAC <- GAC %>% left_join(TFx %>% filter(grepl("_cfWGS", cfSample)), by=c("cfSample"= "sample_id"))}
  GAC <- GAC %>% mutate(cfSample = cfSample.y) %>% select(-cfSample.y)
  
  #filter df of all tumor mutations to the patient we're looking at, and add on tumor purity for each tumor sample
  tumor_sub <- tumor_filtered %>% filter(patient == pat) %>% left_join(tumor_purity %>% select(sample_id, tumor_purity), by=c("sample"="sample_id"))
  
  #left join the cfDNA read counts to the tumor mutations by varID
  df1 <- tumor_sub %>% select(varID, id, Func.refGene, Gene.refGene, sample, patient, Total_reads_in_Tumor, Tumor_VAF, tumor_purity) %>%
    left_join(GAC %>% select(varID, Ref_reads_ctDNA, Alt_reads_ctDNA, Total_reads_ctDNA,cfSample, cfDNA_TFx, type))
  
  #calculate power per tumor mutation using purity corrected tumor VAF and dbinom of cfDNA read depth
  df1 <- df1 %>% 
    mutate(p=Tumor_VAF*cfDNA_TFx/tumor_purity) %>%
    mutate(p_alt = ifelse(p>1, 1, p)) %>%
    mutate(power = 1- (dbinom(0,Total_reads_ctDNA, p_alt)+dbinom(1,Total_reads_ctDNA, p_alt)+dbinom(2,Total_reads_ctDNA, p_alt)))
  allpower <- bind_rows(allpower, df1)
}

allpower <- allpower %>%
  mutate(CAC3 = case_when(Alt_reads_ctDNA>=3~"CAC_called_3", .default = "CAC_under_3"))

#cfDNAPower_TITANTFx_BEDfiltered_TumorMuts.csv: using only 1 TFx per patient, prioritizing TITAN
#write.csv(allpower, "/cfDNA/withTURBT/cfDNAPower_TITANTFx_pycloneBEDfiltered_TumorMuts_withTURBT.csv", row.names=F)
```

### ANNOTATE + COLLAPSE ###

annotate mutation clonality + driver info
```{r}
#grab per mutation cluster & clonality annotation (csv created in Metastatic_seeding/cluster_CPs.Rmd)
mut_clonality <- read.csv('/Supplementary_Tables/genomic_results/SNVs/all_mut_CP_withAnnovar_withTURBT.csv') %>%
  mutate(sample_id= case_match(sample_id, "16-079_Primary-FFPE"~"16-079_Primary-FF", .default = sample_id)) %>%
  rowwise() %>% mutate(id = paste(Chr, Start, Ref, Alt, sep =  "_"), varID = paste(Chr, Start, Ref, sep = "_")) %>% ungroup()

#add tumor cluster annotation to cfDNA mutation calls
allpower <- allpower %>% 
  left_join(mut_clonality %>% select(sample_id, cluster_id, id, cluster_mut_no, cluster_type, pathogenic_3m_updated, IsDriverBLCAupdated), by=c("sample"="sample_id", "id"="id")) %>%
  rename("cluster_status"="cluster_type")

#how many unannotated cluster_status in allpower?
nrow(filter(allpower, is.na(cluster_status))) #79297 rows (not unique mutations)
filter(allpower, is.na(cluster_status)) %>% pull(sample) %>% table() %>% as.data.frame() %>% arrange(desc(Freq)) #18-101, 17-047, 19-001, 19-037 all have >2000
allpower %>% group_by(patient, id) %>% slice_head() %>% filter(is.na(cluster_status)) %>% nrow() #30,085 unique mutations unannotated (of 423,243)
allpower %>% group_by(patient, id) %>% slice_head() %>% filter(is.na(cluster_status)) %>% pull(patient) %>% table() %>% as.data.frame() %>% arrange(desc(Freq))

#write.csv(allpower, "/cfDNA/withTURBT/cfDNAPower_TITANTFx_pycloneBEDfiltered_TumorMuts_withTURBT.csv", row.names=F)
```

collapse to unique mutation per patient 
-powered = power > 0.8 from any tumor
```{r}
collapsed_power <- allpower %>%
  group_by(cfSample, id) %>%
  summarise(powered = any(power>0.8), 
            across(c(patient, Func.refGene, Gene.refGene, cfDNA_TFx, type,CAC3, cluster_id, cluster_mut_no, cluster_status, Ref_reads_ctDNA, Alt_reads_ctDNA, Total_reads_ctDNA, pathogenic_3m_updated, IsDriverBLCAupdated), unique),
            .groups = "drop")

#write.csv(collapsed_power, "/cfDNA/withTURBT/cfDNAPower_TITANTFx_pycloneBEDfiltered_TumorMuts_withTURBT_Collapsed.csv", row.names=F)
```

### Subset TP to BED file w/ 150bp pad ##

import TP BED file (with 150bp buffer)
```{r}
TPbed_hg38_thomas150 <- import.bed("/cell_free_dna/targeted_panel/BroadPanCancer2019.bed.cp.hg38.buffer.txt")
```

which tumor mutations could be captured by TP BED
```{r}
#all tumor mutations
cfTP <- collapsed_power %>% filter(type=="ULP-WGS") %>%
  select(cfSample, id, Gene.refGene) %>%
  mutate(Chr = sub("_.*", "", id), 
         Start = as.numeric(sub(".*_(\\d+)_.*", "\\1", id)), 
         End = as.numeric(sub(".*_(\\d+)_.*", "\\1", id)))

#make into GRanges
G_cfTP <- GRanges(seqnames = cfTP$Chr, 
                            ranges = IRanges(start=cfTP$Start, end=cfTP$End), 
                            name=cfTP$id, 
                            Gene = cfTP$Gene.refGene, 
                            sample = cfTP$cfSample)

#find overlaps between mutations and 150bp buffer TP BED file
cfTP_inBED <- as.data.frame(subsetByOverlaps(G_cfTP, TPbed_hg38_thomas150)) #no buffer = 718 tumor mutations in TP BED file range; thomas 150buffer = 736 in BED range

#add column to collapsed_power with whether the tumor mutation is covered by the targeted panel
inBED <- cfTP_inBED %>% pull(name)
collapsed_power <- collapsed_power %>% 
  mutate(in_TPBED = case_when(type == "cfWGS"~"Yes", 
                              id %in% inBED~"Yes", .default = "No")) %>%
  mutate(type = case_match(type, "ULP-WGS"~"Targeted", "cfWGS"~"WGS"))
```

write out with 150bp BED file column
```{r}
write.csv(collapsed_power, "/cfDNA/withTURBT/cfDNAPower_TITANTFx_pycloneBEDfiltered_TumorMuts_withTURBT_Collapsed_TPbed.csv", row.names = F)
```